Constraint Grammar as a SAT problem

### 2015-05-31 paper/nodalida2015.tex
more text

\noindent The most important differences between the traditional and the SAT-based approach are summarised below:

\paragraph{Condition of being unambiguously tagged is irrelevant.}
Rather than waiting for a word to get disambiguated, the SAT solver starts by 
making assumptions (e.g. ``\emph{casa} is a noun'') and working under them,
discarding the assumption if it doesn't lead to a model that satisfies
all constraints.

The traditional CG relies on that applying the rule set iteratively,
some rules fire the first time and make way for new rules to fire for
the second time.
For SAT-based CG, we present the rules as implications:
``if the \texttt{n} reading for \emph{casa} is true, then discard the
\texttt{prn} reading''. Each rule generates a list these implications,
and the SAT solver is left with the task of finding a model that will
satisfy them.

Let us continue with the earlier example. 
We can add a word to the sentence, \emph{la casa grande} `the big house',
and a rule that removes verb if followed by an adjective.
\begin{itemize}
\item [] \texttt{REMOVE v IF (1 adj) ;}
\end{itemize}
Then the additional clauses will make it impossible for \emph{casa} to
be a verb\footnote{If \texttt{adj} is the only reading for
\emph{grande}, it must be true, because of the restriction that at
least one analysis for each word is true. Then the implication
\texttt{grande<adj> => $\neg$casa<v>} has a true antecedent, thus
its consequent will hold.}, so we will get a unique solution with \texttt{det n adj}.
Standard CG will also remove the verb reading, 
but remember the beginning: if the \texttt{det} reading of \emph{la}
was removed first, then we are stuck with that choice.

@@

\paragraph{Rules can be unordered.}

As hinted by the previous property, SAT solver doesn't need a fixed
order of the rules.
Applying a rule to a sentence produces a number of clauses,
and those clauses are fed into the SAT solver.
However, in the unordered scheme, some information is lost: the
following rule sets would be treated identically, whereas in the
traditional CG, only the first would be considered as a bad order.

\begin{itemize}
\item [\texttt{1)}] \texttt{SELECT verb ;} \\
         \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} \\

\begin{itemize}
\item [\texttt{2)}] \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} \\
         \texttt{SELECT verb ;}
\end{itemize}

@@

\paragraph{Rules disambiguate more.}
Considering our example phrase and rules, standard CG implementation
can only remove readings for the target word (\texttt{prn} or
\texttt{det}). The SAT-based implementation interprets the rules as
``determiner and verb together are illegal'', and is free to take action that concerns also the word in the condition (\texttt{1 n} or \texttt{1 v}).
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. A rule with more conditions corresponds to many
rules, each condition taking its turn to be the target of removal or selection. \\

@@

The way we implement SAT-CG is ideal for this kind of grammar
analysis. After applying each rule, it stores the clauses produced by
it and commits to them -- this makes the execution slow, but it
retains the dependencies.
Given the design that each rule application (instead of rule) gets a variable,
in case of a conflict, the program prints out the particular
application that violates the previous clauses, with the sentence
where it is applied. 
This conflict solving can be done per section, or for the whole grammar. 
In case of section-based, we run each section on the input, and give
the output as an input to the new section, discarding all clauses.

Maximisation-based conflict solving (conflicting rule set --> remove least
possible amount to get a non-conflicting rule set) is not feasible for this task. 
The whole definition of conflicts depends on ordering, and it would
not differentiate between the case where \texttt{SELECT verb} comes
before \texttt{REMOVE verb IF (-1 det)} or after.
On the other hand, a formalism such as FSIG would benefit from the
maximisation-based technique in conflict handling.

@@

% \section*{Acknowledgments}

>>>>

\noindent The most important differences between the traditional and
the SAT-based approach are described in the following sections.

\subsection{Rules disambiguate more}
Considering our example phrase and rules, the standard CG implementation
can only remove readings from the target word (\texttt{prn} or
\texttt{det}). The SAT-based implementation interprets the rules as
``determiner and verb together are illegal'', and is free to take action that concerns also the word in the condition (\texttt{n} or \texttt{v}).

This behaviour is explained by simple properties of logical formulae.
When the rules are applied to the text, they are translated into
implications: \texttt{REMOVE prn IF (1 n)} becomes
 \emph{casa}\texttt{<n>} $\Rightarrow$ $\neg$\emph{la}\texttt{<prn>},
 which reads ``if the \texttt{n} reading for \emph{casa} is true, then
 discard the \texttt{prn} reading for \emph{la}''.
Any implication $a\,\Rightarrow\,b$ can be represented as a disjunction
$\neg a\,\vee\,b$; intuitively, either the antecedent is false
and the consequent can be anything, or the consequent is true and the
antecedent can be anything.
Due to this property, our rule translates into the disjunction 
$\neg$\emph{casa}\texttt{<n>} $\vee$ $\neg$\emph{la}\texttt{<prn>}.

However, our rule is not the only way to reach this form: since
disjunction doesn't care about order, we notice that also
\emph{la}\texttt{<prn>} $\Rightarrow$ $\neg$\emph{casa}\texttt{<n>} 
corresponds to the same disjunction.
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. 
A rule with more conditions corresponds to many
rules, each condition taking its turn to be the target of removal or
selection.

\subsection{Cautious context is irrelevant}
% Rather than waiting for a word to get disambiguated, the SAT solver starts by 
% making assumptions (e.g. ``\emph{casa} is a noun'') and working under them,
% discarding the assumption if it doesn't lead to a model that satisfies
% all constraints.

The traditional CG relies on that applying the rule set iteratively:
some rules fire during the first iteration, either because their
conditions don't require cautious context, or because some words are
unambiguous to start with. This makes some more words unambiguous, and
new rules can fire during the second iteration.

In SAT-CG, the notion of cautious context is irrelevant.
When the rules are applied to the text, they are translated into
implications, such as \\ 
 \emph{casa}\texttt{<n>} $\Rightarrow$ $\neg$\emph{la}\texttt{<prn>},
 which reads ``if the \texttt{n} reading for \emph{casa} is true, then
 discard the \texttt{prn} reading for \emph{la}''. Each rule generates
 a number of these implications, and the SAT solver tries to find a model that will satisfy them.

Let us continue with the earlier example. 
We can add a word to the input:
\begin{itemize}
\item [] \emph{la casa grande} `the big house'
\end{itemize}
and a rule that removes verb reading, if the word is followed by an adjective:
\begin{itemize}
\item [] \texttt{REMOVE v IF (1 adj) ;}
\end{itemize}

% The SAT solver has already the clauses produced by the previous rules.
% \emph{casa}\texttt{<n>} $\Rightarrow$ $\neg$\emph{la}\texttt{<prn>}
% and   \emph{casa}\texttt{<v>} $\Rightarrow$ $\neg$\emph{la}\texttt{<det>}
The new rule adds the implication
 \emph{grande}\texttt{<adj>} $\Rightarrow$
 $\neg$\emph{casa}\texttt{<v>}, which will
disambiguate  \emph{casa} to a noun\footnote{Assuming that
  \texttt{adj} is the only reading for \emph{grande}, it must be true,
  because of the restriction that at least one analysis for each word
  is true. Then the implication has a true antecedent (\emph{grande}\texttt{<adj>}), thus its
  consequent ($\neg$\emph{casa}\texttt{<v>}) will hold.}. 
With the status of \emph{casa} resolved, the SAT solver can now make
decision between the two models that it would find with only \emph{la casa}.
Now \emph{casa}\texttt{<n>} is true, so \emph{la}\texttt{<prn>} must
be false. 
Because \emph{la}\texttt{<prn>} is false,  \emph{casa} cannot
 be a \texttt{v}, and we get a unique solution with \texttt{det n adj}.

Contrast this with the behaviour of the standard CG.

With the new rule, standard CG will also remove the verb reading from \emph{casa}, 
but it is in no way connected to the choice for \emph{la}. It all
depends of the order of the two rules; if the \texttt{det} reading of \emph{la} is removed first, then we are stuck with that choice.

@@

\subsection{Rules can be unordered}

As hinted by the previous property, SAT solver doesn't need a fixed
order of the rules.
Applying a rule to a sentence produces a number of clauses,
and those clauses are fed into the SAT solver.
However, in the unordered scheme, some information is lost: the
following rule sets would be treated identically, whereas in the
traditional CG, only the second would be considered as a bad order.

\begin{itemize}
\item [\texttt{1)}] \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} \\
         \texttt{SELECT verb ;}
\item [\texttt{2)}] \texttt{SELECT verb ;} \\
         \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} 
\end{itemize}

@@

@@

The SAT-CG implementation is ideal for this kind of corpus-based grammar
analysis. After applying each rule, it stores the clauses produced by
it and commits to them -- this makes the execution slow, but it
retains the dependencies.
Given the design that each rule application (instead of rule) gets a variable,
in case of a conflict, the program prints out the particular
application that violates the previous clauses, with the sentence
where it is applied. 
This conflict solving can be done per section, or for the whole grammar. 
In case of section-based, we run each section on the input, and give
the output as an input to the new section, discarding all clauses.

Maximisation-based conflict solving (see Section 3) is not feasible for this task. 
The whole definition of conflicts depends on ordering, and it would
not differentiate between the case where \texttt{SELECT verb} comes
before \texttt{REMOVE verb IF (-1 det)} or after.
On the other hand, a formalism such as FSIG would benefit from the
maximisation-based technique in conflict handling.

For future work, we intend to apply the SAT solver in testing for
conflicts without a corpus. Let us illustrate the idea with the rule
set shown in Section 3:

\begin{itemize}
\item [\texttt{1)}] \texttt{SELECT verb ;} \\
         \texttt{REMOVE verb IF (-1 det) ;} \\

\begin{itemize}
\item [\texttt{2)}] \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{SELECT verb ;}
\end{itemize}

Assume we have tag set \texttt{\{det, noun, verb, sg, pl\}}. Then we
can ask the SAT solver a question if there exists an input which
satisfies our constraints: both rules, applied in the order, remove
something from the input.
after applying the first rule, can be modified by the second rule.

Implementing the design on a larger scale is left for future work.

@@

\section*{Acknowledgments}

We thank Tino Didriksen, Eckhard Bick, Francis Tyers and Anssi
Yli-Jyrä for comments and suggestions; also everyone at the CG
workshop who participated in the discussion.
